---
title: "07.2-DML-adult-comparison-low-cov"
author: "Kathleen Durkin"
date: "2025-06-10"
always_allow_html: true
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

In `06.2-diff-methyl-split-lifestage-low-cov` I identified loci in the larval samples that are differentially methylated based on parental OA treatment. I now want to see whether any of those DMLs were also present in the adults

Note: I'm first checking the adult DML tables, but I may also want to look a the methylomes of both the larvae and the larvae's parents in more detail, since not all adult oysters whose gametes were sequenced in Venkatamaran et al. 2024 were used for fertilization crosses.

Download adult DMLs (both female (eggs) and male (sperm))
```{r, engine='bash', eval=FALSE}
curl -L https://github.com/sr320/ceabigr/raw/refs/heads/main/data/female_dml.bed -o ../data/adult_female_dml.bed
curl -L https://github.com/sr320/ceabigr/raw/refs/heads/main/data/male_dml.bed -o ../data/adult_male_dml.bed
```

## Combo

```{r, engine='bash'}
echo "Combo DMLs that are also differentially methylated in female parents"
/home/shared/bedtools2/bin/bedtools intersect -a ../output/06.2-diff-methyl-split-lifestage-low-cov/combo_diff25_DML.bed -b ../data/adult_female_dml.bed

echo ""

echo "Combo DMLs that are also differentially methylated in male parents"
/home/shared/bedtools2/bin/bedtools intersect -a ../output/06.2-diff-methyl-split-lifestage-low-cov/combo_diff25_DML.bed -b ../data/adult_male_dml.bed

echo""

echo "Total female:"
/home/shared/bedtools2/bin/bedtools intersect -a ../output/06.2-diff-methyl-split-lifestage-low-cov/combo_diff25_DML.bed -b ../data/adult_female_dml.bed | wc -l
echo ""
echo "Total male:"
/home/shared/bedtools2/bin/bedtools intersect -a ../output/06.2-diff-methyl-split-lifestage-low-cov/combo_diff25_DML.bed -b ../data/adult_male_dml.bed | wc -l
```

7 Combo DMLs present in the male parents (0 in the female)

## Zygote

```{r, engine='bash'}
echo "Larval DMLs that are also differentially methylated in female parents"
/home/shared/bedtools2/bin/bedtools intersect -a ../output/06.2-diff-methyl-split-lifestage-low-cov/zygote_diff25_DML.bed -b ../data/adult_female_dml.bed

echo ""

echo "Larval DMLs that are also differentially methylated in male parents"
/home/shared/bedtools2/bin/bedtools intersect -a ../output/06.2-diff-methyl-split-lifestage-low-cov/zygote_diff25_DML.bed -b ../data/adult_male_dml.bed

echo""

echo "Total female:"
/home/shared/bedtools2/bin/bedtools intersect -a ../output/06.2-diff-methyl-split-lifestage-low-cov/zygote_diff25_DML.bed -b ../data/adult_female_dml.bed | wc -l
echo ""
echo "Total male:"
/home/shared/bedtools2/bin/bedtools intersect -a ../output/06.2-diff-methyl-split-lifestage-low-cov/zygote_diff25_DML.bed -b ../data/adult_male_dml.bed | wc -l
```

15 Zygote DMLs present in the male parents (0 in the female)

## Larvae

```{r, engine='bash'}
echo "Larval DMLs that are also differentially methylated in female parents"
/home/shared/bedtools2/bin/bedtools intersect -a ../output/06.2-diff-methyl-split-lifestage-low-cov/larvae_diff25_DML.bed -b ../data/adult_female_dml.bed

echo ""

echo "Larval DMLs that are also differentially methylated in male parents"
/home/shared/bedtools2/bin/bedtools intersect -a ../output/06.2-diff-methyl-split-lifestage-low-cov/larvae_diff25_DML.bed -b ../data/adult_male_dml.bed

echo""

echo "Total female:"
/home/shared/bedtools2/bin/bedtools intersect -a ../output/06.2-diff-methyl-split-lifestage-low-cov/larvae_diff25_DML.bed -b ../data/adult_female_dml.bed | wc -l
echo ""
echo "Total male:"
/home/shared/bedtools2/bin/bedtools intersect -a ../output/06.2-diff-methyl-split-lifestage-low-cov/larvae_diff25_DML.bed -b ../data/adult_male_dml.bed | wc -l
```

31 Larvae DMLs present in male parents, 2 in the female parents (though both have nearly insignificant methylation differences)
