---
title: "09-multifactor-diff-methyl"
author: "Kathleen Durkin"
date: "2025-06-10"
always_allow_html: true
output: 
  github_document:
  toc: true
toc_depth: 3
number_sections: true
html_preview: true 
bookdown::html_document2:
  theme: cosmo
toc: true
toc_float: true
code_folding: show
code_download: true
---

Again, I'll be excluding the following low-quantity samples

- CF01-CM01-Zygote (40.4 M)

- CF08-CM04-Larvae (6.2 M)

- CF08-CM05-Larvae (6.2 M)

- EF04-EM04-Zygote (97.3 M)

- EF05-EM05-Zygote (53.8 M)


Inputs: 

-   Methylation calls extracted from WGBS reads using `Bismark`: `[].fastp-trim_bismark_bt2_pe.deduplicated.bismark.cov.gz`, stored on Gannet: [offspring data](https://gannet.fish.washington.edu/gitrepos/ceasmallr/output/02.20-bismark-methylation-extraction/), [parent data](https://gannet.fish.washington.edu/seashell/bu-mox/scrubbed/120321-cvBS/)


# Download methylation calls if needed

Offspring:
```{r, engine='bash', eval=FALSE}
# Run wget to retrieve FastQs and MD5 files
# Note: the --no-clobber command will skip re-downloading any files that are already present in the output directory
wget \
--directory-prefix ../data/bismark-methyl-extraction \
--recursive \
--no-check-certificate \
--continue \
--cut-dirs 4 \
--no-host-directories \
--no-parent \
--quiet \
--no-clobber \
--accept="*.deduplicated.bismark.cov.gz,checksums.md5" https://gannet.fish.washington.edu/gitrepos/ceasmallr/output/02.20-bismark-methylation-extraction/

```

```{r, engine='bash', eval=FALSE}
cd ../data/bismark-methyl-extraction/

echo "How many methylation calling files were downloaded?"
ls *.deduplicated.bismark.cov.gz | wc -l

echo ""

echo "Check file checksums:"
grep '.deduplicated.bismark.cov.gz' checksums.md5 | md5sum -c -
```

We should have N=32 (n=15 ControlxControl crosses, n=17 ExposedxExposed crosses)

Parent:
```{r, engine='bash'}
# Parent metadata
curl -L https://github.com/sr320/ceabigr/raw/refs/heads/main/data/adult-meta.csv -o ../data/adult-meta.csv
```

```{r, engine='bash', eval=FALSE}
# Run wget to retrieve FastQs and MD5 files
# Note: the --no-clobber command will skip re-downloading any files that are already present in the output directory
wget \
--directory-prefix ../data/bismark-methyl-extraction-parent \
--recursive \
--no-check-certificate \
--continue \
--cut-dirs 4 \
--no-host-directories \
--no-parent \
--quiet \
--no-clobber \
--accept="*.deduplicated.bismark.cov.gz,checksums.md5" https://gannet.fish.washington.edu/seashell/bu-mox/scrubbed/120321-cvBS/

```

Rename files to instead use the updated Sample IDs
```{r, engine='bash'}
cd ../data/bismark-methyl-extraction-parent/

metadata="../adult-meta.csv"

# Skip header and read line by line
tail -n +2 "$metadata" | while IFS=',' read -r SampleID OldSampleID Treatment Sex TreatmentN ParentID; do
  # Find files starting with SampleID
  for file in ${SampleID}*; do
    # Check that file exists to avoid glob expansion issues
    [ -e "$file" ] || continue

    # New file name: replace SampleID with ParentID
    newname="${file/$SampleID/$ParentID}"

    # Rename the file
    echo "Renaming $file to $newname"
    mv "$file" "$newname"
  done
done
```

# methylKit

Load packages
```{r}
# Install packages if necessary
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("methylKit")
# BiocManager::install("genomation")

# Load
library(tidyr)
library(dplyr)
library(methylKit)
library(genomation)
library(GenomicRanges)
library(gridExtra)
library(ggplot2)
library(ggfortify)
```

# Offspring

## Create object

```{r, cache=FALSE}

# Define path to my samples directory
data_dir_offspring <- "../data/bismark-methyl-extraction"

# List all .bismark.cov.gz files in the samples directory
file.list_offspring <- list.files(path = data_dir_offspring, pattern = "\\.bismark.cov.gz$", full.names = TRUE)
# Exclude samples with less than 100M cytosines analyzed (poor coverage)
file.list_offspring <- file.list_offspring[!grepl("CF01-CM01-Zygote|CF08-CM04-Larvae|CF08-CM05-Larvae|EF04-EM04-Zygote|EF05-EM05-Zygote", file.list_offspring)]


# Extract just the filenames
file.names_offspring <- basename(file.list_offspring)

# Create sample IDs by stripping file extensions
sample.ids_offspring <- sub("_R1_001.fastp-trim_bismark_bt2_pe.deduplicated.bismark.cov.gz", "", file.names_offspring)
sample.ids_offspring <- sub("_R1_001.fastp-trim.REPAIRED_bismark_bt2_pe.deduplicated.bismark.cov.gz", ".REP", sample.ids_offspring)


# Assign treatment based on first letter of filename
# E = parents exposed to OA treatment
# C = parents held in control
treatment_offspring <- ifelse(startsWith(file.names_offspring, "E"), 1,
                    ifelse(startsWith(file.names_offspring, "C"), 0, NA))

# Read in methylation data
myobj_offspring <- methRead(as.list(file.list_offspring),
  sample.id = as.list(sample.ids_offspring),
  pipeline = "bismarkCoverage",
  assembly = "mm10",
  treatment = treatment_offspring,
  mincov = 5
)

# Inspect
str(myobj_offspring)
head(myobj_offspring[[1]])
```

## Filter

We want to filter out bases with low coverage (<10 reads) because they will reduce reliability in downstream analyses. We also want to discard any bases with exceptionally high coverage (in the 99.9th percentile), as this is an indication of PCR amplification bias.
```{r}
myobj.filt_offspring <- filterByCoverage(myobj_offspring,
                      lo.count=5,
                      lo.perc=NULL,
                      hi.count=NULL,
                      hi.perc=99.9)
```

## Normalization

Next, we normalize the coverage values among samples.

```{r}
myobj.filt.norm_offspring <- normalizeCoverage(myobj.filt_offspring, method = "median")
```

## Merge

Data doesn't appear to be two-stranded, so will use `destrand=FALSE`

```{r, cache=TRUE}
methylBase_offspring <- methylKit::unite(myobj.filt.norm_offspring, destrand=FALSE)
nrow(methylBase_offspring)
```

After removing low-quantity samples and reducing minimum coverage, we retain 22,874 CpG sites


# Parents

## Create object

```{r, cache=FALSE}

# Define path to my samples directory
data_dir_parents <- "../data/bismark-methyl-extraction-parent"

# List all .bismark.cov.gz files in the samples directory
file.list_parents <- list.files(path = data_dir_parents, pattern = "\\.bismark.cov.gz$", full.names = TRUE)

# Extract just the filenames
file.names_parents <- basename(file.list_parents)

# Create sample IDs by stripping file extensions
sample.ids_parents <- sub("_R1_val_1_bismark_bt2_pe.deduplicated.bismark.cov", "", file.names_parents)


# Assign treatment based on first letter of filename
# E = parents exposed to OA treatment
# C = parents held in control
treatment_parents <- ifelse(startsWith(file.names_parents, "E"), 1,
                    ifelse(startsWith(file.names_parents, "C"), 0, NA))

# Read in methylation data
myobj_parents <- methRead(as.list(file.list_parents),
  sample.id = as.list(sample.ids_parents),
  pipeline = "bismarkCoverage",
  assembly = "mm10",
  treatment = treatment_parents,
  mincov = 5
)

# Inspect
str(myobj_parents)
head(myobj_parents[[1]])
```

## Filter

We want to filter out bases with low coverage (<10 reads) because they will reduce reliability in downstream analyses. We also want to discard any bases with exceptionally high coverage (in the 99.9th percentile), as this is an indication of PCR amplification bias.
```{r}
myobj.filt_parents <- filterByCoverage(myobj_parents,
                      lo.count=10,
                      lo.perc=NULL,
                      hi.count=NULL,
                      hi.perc=99.9)
```

## Normalization

Next, we normalize the coverage values among samples.
```{r}
myobj.filt.norm_parents <- normalizeCoverage(myobj.filt_parents, method = "median")
```

## Merge

Data doesn't appear to be two-stranded, so will use `destrand=FALSE`
```{r, cache=TRUE}
methylBase_parents <- methylKit::unite(myobj.filt.norm_parents, destrand=FALSE)
nrow(methylBase_parents)
```

After removing low-quantity samples and reducing minimum coverage, we retain 2,447,226 CpG sites


# PCA

```{r}
# Extract percent methylation matrices
meth_offspring <- percMethylation(methylBase_offspring)
meth_parent    <- percMethylation(methylBase_parents)

# Only keep positions that are present in all offspring *and* parents
# Get positions for each matrix (chr, start, strand)
pos_offspring <- getData(methylBase_offspring)[, c("chr", "start", "strand")]
pos_parent    <- getData(methylBase_parents)[, c("chr", "start", "strand")]

# Find shared CpGs
common_pos <- inner_join(pos_offspring, pos_parent, by = c("chr", "start", "strand"))

# Subset matrices to shared positions
# Create row indices for shared CpGs in both matrices
idx_offspring <- which(paste(pos_offspring$chr, pos_offspring$start, pos_offspring$strand) %in%
                         paste(common_pos$chr, common_pos$start, common_pos$strand))
idx_parent <- which(paste(pos_parent$chr, pos_parent$start, pos_parent$strand) %in%
                      paste(common_pos$chr, common_pos$start, common_pos$strand))

meth_offspring_common <- meth_offspring[idx_offspring, ]
meth_parent_common    <- meth_parent[idx_parent, ]

# Combine the matrices
combined_matrix <- cbind(meth_offspring_common, meth_parent_common)

# Transpose for PCA
pca_input <- t(combined_matrix)

# Remove CpGs (columns) with zero variance across all samples (will break PCA)
pca_input_filtered <- pca_input[, apply(pca_input, 2, var, na.rm = TRUE) != 0]

# Run PCA
pca_result <- prcomp(pca_input_filtered, scale. = TRUE)
```

```{r}
# Create metadata for plotting
sample_names <- rownames(pca_input_filtered)
# Identify sample roles and traits
metadata <- data.frame(
  Sample = sample_names,
  stringsAsFactors = FALSE
) %>%
  mutate(
    # Assign 'Offspring' if name contains a dash (e.g., CF01-CM03)
    Role = ifelse(grepl("-", Sample), "Offspring", "Parent"),
    
    # Extract treatment from the first character
    Treatment = substr(Sample, 1, 1),
    
    # Assign combined sex/role label
    Sex = case_when(
      Role == "Offspring" ~ "Offspring",
      grepl("M", substr(Sample, 2, 2)) ~ "Male Parent",
      grepl("F", substr(Sample, 2, 2)) ~ "Female Parent",
      TRUE ~ "Unknown"
    ),
    
    # Clean up treatment
    Treatment = recode(Treatment, "C" = "Control", "E" = "Exposed")
  )

autoplot(pca_result, data = metadata, colour = "Sex", shape = "Treatment", size=2) +
  ggtitle("PCA of Methylation Profiles") +
  theme_minimal() +
  theme(legend.position = "right")
```


